---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# The Groundwater Monitoring Infrastructure Assessment Dataset, Nkhotakota - Malawi (2022)

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/1326500897.svg)](https://doi.org/10.5281/zenodo.21839342)
[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

The Groundwater Monitoring Infrastructure Assessment Dataset (2022) is
an open dataset published by BASEflow Malawi. The dataset contains field
observations collected during an assessment of groundwater monitoring
infrastructure in Nkhotakota District, Malawi in 2022. It includes
information on monitoring well characteristics, groundwater level
measurements, monitoring equipment, site accessibility, infrastructure
condition, community awareness, and recommended maintenance and
rehabilitation actions.

The dataset captures key attributes of groundwater monitoring sites,
including well depth, static groundwater level, measurement methods,
level logger and barometric logger availability, equipment condition,
site accessibility, monitoring well purpose, vandalism risks, nearby
water source types, and infrastructure improvement recommendations.
Together, these data provide a comprehensive assessment of the
operational status and maintenance needs of groundwater monitoring
infrastructure in the study area.

This dataset supports evidence-based groundwater management and can be
used to evaluate the condition of monitoring infrastructure, identify
maintenance priorities, strengthen groundwater monitoring networks, and
support sustainable groundwater resource management in Malawi.

**Potential Use Cases**

The dataset is suitable for use by:

1.  Government ministries and water authorities responsible for
    groundwater monitoring and water resources management.

2.  Researchers and universities conducting studies in hydrogeology,
    groundwater monitoring, climate resilience, and environmental
    management.

3.  Development partners and non-governmental organizations implementing
    groundwater and water security programmes.

4.  Hydrogeologists and groundwater professionals assessing monitoring
    network performance and infrastructure condition.

5.  GIS analysts and data scientists developing spatial analyses,
    dashboards, and predictive models for groundwater monitoring.

6.  Policy makers supporting groundwater governance, infrastructure
    planning, and investment decisions.

7.  Students undertaking research in groundwater, environmental science,
    civil engineering, and water resources.

8.  Organizations contributing to Sustainable Development Goal 6 (Clean
    Water and Sanitation), integrated water resources management, and
    climate adaptation initiatives.

## Installation

You can install the development version of gwinfraassess from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/gwinfraassess")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/gwinfraassess/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to The Groundwater Monitoring Infrastructure Assessment Dataset collected by BASEflow in 2022, in Nkhotakota district.

```{r}
library(gwinfraassess)
```

### gwinfraassess

The dataset `gwinfraassess` contains 
`r nrow(gwinfraassess)` observations and `r ncol(gwinfraassess)`
variables

```{r}
gwinfraassess |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "gwinfraassess.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
library(gwinfraassess)

# Visualization: Improvement Recommendations
# libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

#filter rows where improvement_plan is NA or null
improve_plot <-
  gwinfraassess %>%
  filter(!is.na(improvement_plan),
         improvement_plan != "") %>%
  separate_rows(improvement_plan, sep = ",") %>%
  mutate(improvement_plan = str_trim(improvement_plan)) %>%
  count(improvement_plan, sort = TRUE)
#Plot the results
ggplot(improve_plot,
       aes(x = reorder(improvement_plan, n),
           y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Recommended Improvement Actions",
    x = NULL,
    y = "Number of Monitoring Wells"
  ) +
  theme_minimal(base_size = 12)
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("gwinfraassess")
```
